Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-05-06 exp1/iswc2013.tex
large dataset explanation
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@517 43ef1190-90fd-4dd7-a769-d0121f02b80e

The resulting normal plot is in \figure{fig:frac32normal}. We see that
only two main effects are significant on a $\alpha = 0.05$ level
according to the Lenth criterion, ``Implement'' is again deemed a
significant improvement, and ``Union'' is deemed significantly the
hardest operation for the query engine. We see that some other points
also deviate from the straight line, but with as few runs as this, the
total variance is great, so no other conclusions can be drawn. Also,
the Student's t-test we used in the previous section cannot be
employed, as we have no additional data to use for the test.

For any other purpose than a rough idea of some key influencers on the
overall performance of the endpoint, this little experiment is insufficient.

@@

>>>>

The resulting normal plot is in figure~\ref{fig:frac32normal}. We see that
only two main effects are significant on a $\alpha = 0.05$ level
according to the Lenth criterion, ``Implement'' is again deemed a
significant improvement, and ``Union'' is deemed significantly the
hardest operation for the query engine. We see that some other points
also deviate from the straight line, but with as few runs as this, the
total variance is great, so no other conclusions can be drawn. Also,
the Student's t-test we used in the previous section cannot be
employed, as we have no additional data to use for the test.

For any other purpose than a rough idea of some key influencers on the
overall performance of the endpoint for the given levels, this little
experiment is insufficient.

We turn to the 64-run experiment and its normal plot in
figure~\ref{fig:frac64normalalpha005}. We see that the we get more significant
effects, and that they correspond well to those we saw in the full
factorial experiment. They also all have an intuitive
explanation. However, we do not see the effects that we found
worrisome in the full factorial design. The ``Implement:TripleC''
interaction emerges only for a level $\alpha=0.15$, see
figure~\ref{fig:frac64normalalpha015}. To ensure that we discover such
cases will be a key challenge in further studies.

\todo{Do t-tests}

@@

The importance of finding flaws in the levels is highlighted by the
normal plot in figure~\ref{fig:frac64hugenormal}. The smaller
experiment made it feasible to run with a larger dataset, this
experiment uses the original 15~MTriples dataset of
\cite{mxro:Morsey2011DBpedia}. The picture is now as hinted in the
previous discussion, for larger ``TripleC'', our simulated performance
is actually worse, as evident from the positive
``Implement''. We also see that the interactions ``Implement:Lang''
and ``Implement:TripleC'' is highly significant. It is clear what
happened: For larger ``TripleC'', the delay we introduced in the
langMatches function has a devastating effect, since it delays the
execution for every returned triple, and thereby completely dominates
the execution time.

Moreover, we note that the ``Machine'' factor has become highly
significant in the other end along with its ``Machine:TripleC''
interaction. One should take careful notice that the ``Machine''
factor encompasses many different things, that should be different
factors; CPU, disks, available RAM, installed software, tuning
parameters and so on. It is advantegous to have such broad factors in
the design in early experiments, because they may point out unexpected
effects that should be further scrutinized. The ``Machine'' factor is
a good example of usage that also helps us manage
complexity: If possible, one may group many factors into one if one is
in doubt of its relevance and break it up in subsequent
experiments. In this case, the cause of the performance problem was
highly audible: It was the old P-ATA disks of the larger machine, but
in general, such clues are not available to the experimenter.

We are forced to conclude that the three preceeding experiments are
severely flawed, as they failed to take into account the non-linear
effects arising from larger data sizes on slow hardware.

Note that we do not take advantage of the usual performance of 4store
in this simulation, to the contrary, we chose 4store because of its
relative simplicity to modify it to suite our needs for illustrative
purposes, which consists of degrading it in many ways. The small
datasets used here does not imply that the approach will not work for
large datasets; properly configured, actual optimizations will give
very different runtimes, and the small number of runs will make it
possible to evaluate a very large number of factors.
